# R Packages
Before we get started, let's make sure to install and load some necessary R packages. 

## What are packages? 
Packages are a set of codes or scripts that an author has compiled together.   

## Packages
The main packages that we will use throughout the book are:

1. `psych` : Descriptive statistics
2. `car` : Analysis of Variance (ANOVA) source table with sums of squares type III
3. `tidyverse` : Data manipulation and visualization
4. `knitr` : Nice, clean-looking tables
5. `carData` : Open access datasets
6. `MASS` : Additional open access datasets
7. `scales`: Convert numbers into dollars

Note: The description of each package is only a description of how we will use each package. Each package can perform many more functions than I have described. I recommend briefly reading them by typing `??name of package` to read its help page (e.g., `??psych`).

## Install R Packages
To install a package, we can enter the command `install.packages("name of package")`. For example, to install the `psych` package, we would type `install.packages("psych")`. 

```{r, eval=F}
install.packages("psych")
install.packages("car")
install.packages("tidyverse")
install.packages("knitr")
install.packages("kableExtra")
install.packages("carData")
install.packages("MASS")
install.packages("scales")
```

## Load R Packages
To load a package, we can enter the command `library(name of package)`. For example, to load the `psych` package, we would type `library(psych)`. We won't be loading the `carData` and `MASS` package as we will be extracting the datasets directly from those packages, which we will go over in the [datasets chapter][Datasets].

```{r, echo = T, warning = F, message = F}
library(psych)
library(car)
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
```
